Gusto: a toolkit for compatible finite element dynamical cores

### 2025-08-06 source/paper.tex
fix issues from merge

\begin{equation}
  c_p\int_\Omega\MM{w}\cdot\theta\nabla\Pi\diff x = -c_p\int_\Omega\nabla\cdot(\MM{w}\theta)\Pi\diff x + c_p\int_{\Gamma_v}\jump{\theta\MM{w}\cdot\MM{n}}\avg{\Pi} \diff S, \quad \forall \MM{w} \in \ring{\mathbb{V}}^{HDiv},
\end{equation}
where $\Gamma_v$ denotes the vertical element facets, across which
$\theta$, and therefore $\Pi$, are discontinuous. Here we have chosen,
as in \citet{}, to take the average value of $\Pi$ across the facets,
$\avg{\Pi} = (\Pi^++\Pi^-)/2$, where the $+$ and $-$ indicate each
side of the facet. The brackets $\jump{q} = q^+-q^-$ denote the jump
in $q$ across the facet and the integral over the horizontal facets
vanishes since both $\theta$ and $\MM{w}\cdot\MM{n}$ are continuous
across these facets so the jump is zero.

\subsection{Transport terms}
Since the prognostic variables are discretised using different finite
element function spaces $\mathbb{V}^{HDiv}$, $\mathbb{V}^{L^2}$ and
$\mathbb{V}_\theta$, each with different continuity properties between
the elements, the choice of stable transport discretisation is
different for each prognostic variable. The spatial order of the
finite element spaces also imposes requirements on the choice of
transport scheme. In particular, when the lowest order space
$\mathbb{V}^{L2}$ is chosen to be piecewise constant, we need to use
the recovered transport scheme described in
\citet{bendall2019recovered} to achieve second order convergence
overall. Here we describe the default transport schemes used in Gusto
for the next-to-lowest order sequence of function spaces (i.e. where
the lowest order space $\mathbb{V}^{L2}$ is chosen to be piecewise
linear and discontinuous between elements) and refer the reader to
\citet{bendall2019recovered} for details of the recovered schemes,
which are also implemented in Gusto.

\subsubsection{Transport of $\mathbb{V}^{L^2}$ fields}
The shallow water and compressible Euler equation sets both contain a
prognostic equation that is the continuity form of the transport
equation for a variable represented by a function that is fully
discontinuous between elements. This means that the divergence term is
not well defined. A stable discretiation of this equation can be
constructed by integrating that term by parts and taking the upwind
values on the element facets. Using the compressible Euler equation
notation, this gives, in each element $e$ with boundary $\partial e$:

\begin{equation}
  \int_e \phi\nabla\cdot(\rho\MM{u}) \diff x = -\int_e \nabla\phi \cdot (\rho\MM{u}) \diff x + \int_{\partial e} \phi\tilde{\rho}\MM{u}\cdot\MM{n} \diff S, \quad \forall \phi \in \mathbb{V}^{L^2}.
\end{equation}
Summing over the domain gives
\begin{equation}
  \int_\Omega \phi\nabla\cdot(\rho\MM{u}) \diff x = -\int_\Omega \nabla\phi \cdot (\rho\MM{u}) \diff x + \int_\Gamma \jump{\phi\MM{u}}\tilde{\rho} \diff S, \quad \forall \phi \in \mathbb{V}^{L^2},
\end{equation}
where $\Omega$ is the domain, $\Gamma$ denotes the element facets,
$\tilde{\rho}$ denotes the value of $\rho$ on the upwind side of the
facet and the brackets $\jump{q} = q^+-q^-$ denote the jump in $q$
across the facet.

\subsubsection{Transport of $\mathbb{V}_\theta$ fields}
The compressible Boussinesq and compressible Euler equation sets both
contain a prognostic equation that is the advective form of the
transport equation for a variable that is represented by a function
that is continuous in the vertical direction but discontinuous in the
horizontal direction. This means that the gradient of the field
required in the transport term is not well defined. A stable
discretisation of this equation can be constructed by using a
combination of the above upwinding strategy in the discontinuous
horizontal direction and a streamline-upwind Petrov-Galerkin (SUPG)
method in the continuous vertical dirction. The SUPG method replaces
the finite element test function $\gamma$ with $\gamma + \tau
\MM{u}\cdot\nabla\gamma$ where $\tau$ is a stabilisation parameter,
although here we just require the vertical part of the gradient term. This
has the effect of suppressing spurious oscillation associated with the
continuous finite element approximation of the transport term. The
stabilisation parameter $\tau$ depends on the spatial resolution and
** and is set to be ** by default in Gusto, following \citet{}.

We first integrate by parts in each column $C$ of the mesh, taking the
upwind value of the field, indicated by a $\tilde{}$, on the vertical
facets of the mesh, denoted by $\Gamma_v$. Using the notation of the
compressible Euler equations, this gives

\begin{align}
  \int_C \gamma\MM{u}\cdot\nabla\theta\diff x &= -\int_C \nabla\cdot(\gamma\MM{u})\theta \diff x + \int_{\Gamma_v}\gamma \MM{u}\cdot\MM{n}\tilde{\theta}\diff S, \quad \forall \gamma \in \mathbb{V}_\theta, \\
  &= \int_C \gamma\MM{u}\cdot\nabla\theta \diff x + \int_{\Gamma_v}\gamma \MM{u}\cdot\MM{n}(\tilde{\theta}-\theta_{int})\diff S, \quad \forall \gamma \in \mathbb{V}_\theta,
\end{align}
where we have integrated by parts a second time to move the
differential operator from $\gamma$ back to $\theta$ since we need to
replace the test function $\gamma$ with $\gamma+\tau w\gamma_z$, where
$w=\MM{u}\cdot\hat{\MM{z}}$ is the vertical component of the
velocity. This gives a second boundary term that now contains the
value of $\theta$ in the interior of the column, denoted by
$\theta_{int}$. Summing over the domain and applying the SUPG method
in the vertical gives
\begin{equation}
  \int_C \gamma\MM{u}\cdot\nabla\theta\diff x =\int_\Omega(\gamma+\tau w\gamma_z)\MM{u}\cdot\nabla\theta \diff x + \int_{\Gamma_v}\jump{(\gamma+\tau w\gamma_z) \MM{u}}\tilde{\theta}-\jump{(\gamma+\tau w\gamma_z) \MM{u}\theta}\diff S, \quad \forall \gamma \in \mathbb{V}_\theta.
\end{equation}
Note that, as explained in \citet{Colin}, the test function should
also be replaced in the time derivative term for consistency **

\subsubsection{Transport of $\mathbb{V}^{HDiv}$ fields}
All three of the equation sets described in section \ref{} include a
prognostic equation for the velocity field
$\MM{u}\in\mathbb{V}^{HDiv}$ that contains a nonlinear transport term
$\MM{u}\cdot\nabla\MM{u}$. This term can be discretised directly using
a vector version of the scheme in section ** or, as in \citet{} it can
be replaced by the vector invariant form and then discretised. In
Gusto we offer both options and we present the vector invariant scheme
below.

The schemes described above do not preserve monotonicity unlike the
continuous equation, generating spurious over and undershoots. When
monotonicity is important, for example, to avoid generating spurious
and unphysical negative values of moisture variables, we can apply a
limiter as part of the transport scheme. These limiters are described
in \ref{}.

It is common in weather and climate models for the horizontal and
vertical directions to be discretised differently due to the different
scales of motion in each direction. Typically the vertical part of the
transport term is treated implicitly so that the timestep is not
restricted by the vertical Courant number. We provide the option to
split the transport terms into horizontal and vertical components, as
described in \ref{}.

@@

\authorcontribution{TEXT} %% this section is mandatory

>>>>

\begin{equation}
  c_p\int_\Omega\MM{w}\cdot\theta\nabla\Pi\diff x = -c_p\int_\Omega\nabla\cdot(\MM{w}\theta)\Pi\diff x + c_p\int_{\Gamma_v}\jump{\theta\MM{w}\cdot\MM{n}}\avg{\Pi} \diff S, \quad \forall \MM{w} \in \ring{\mathbb{V}}^{HDiv},
\end{equation}
where $\Gamma_v$ denotes the vertical element facets, across which
$\theta$, and therefore $\Pi$, are discontinuous. Here we have chosen,
as in \citet{natale2016compatible}, to take the average value of $\Pi$
across the facets, $\avg{\Pi} = (\Pi^++\Pi^-)/2$, where the $+$ and
$-$ indicate each side of the facet. The brackets $\jump{q} = q^+-q^-$
denote the jump in $q$ across the facet and the integral over the
horizontal facets vanishes since both $\theta$ and $\MM{w}\cdot\MM{n}$
are continuous across these facets so the jump is zero.

\subsection{Transport terms}
\label{sec: transport}
Since the prognostic variables are discretised using different finite
element function spaces $\mathbb{V}^{HDiv}$, $\mathbb{V}^{L^2}$ and
$\mathbb{V}_\theta$, each with different continuity properties between
the elements, the choice of stable transport discretisation is
different for each prognostic variable. The spatial order of the
finite element spaces also imposes requirements on the choice of
transport scheme. In particular, when the lowest order space
$\mathbb{V}^{L^2}$ is chosen to be piecewise constant, we need to use
the recovered transport scheme described in
\citet{bendall2019recovered} to achieve second order convergence
overall. Here we describe the default transport schemes used in Gusto
for the next-to-lowest order sequence of function spaces (i.e. where
the lowest order space $\mathbb{V}^{L^2}$ is chosen to be piecewise
linear and discontinuous between elements) and refer the reader to
\citet{bendall2019recovered} for details of the recovered schemes,
which are also implemented in Gusto.

\subsubsection{Transport of $\mathbb{V}^{L^2}$ fields}
\label{subsec: DG transport}
The shallow water and compressible Euler equation sets both contain a
prognostic equation that is the continuity form of the transport
equation for a variable represented by a function that is fully
discontinuous between elements. This means that the divergence term is
not globally defined. A stable discretisation of this equation can be
constructed by integrating that term by parts and taking the upwind
values on the element facets. Using the compressible Euler equation
notation, this gives, in each element $e$ with boundary $\partial e$:

\begin{equation}
  \int_e \phi\nabla\cdot(\rho\MM{u}) \diff x = -\int_e \nabla\phi \cdot (\rho\MM{u}) \diff x + \int_{\partial e} \phi\hat{\rho}\MM{u}\cdot\MM{n} \diff S, \quad \forall \phi \in \mathbb{V}^{L^2}.
\end{equation}
where $\hat{\rho}$ denotes the value of $\rho$ on the element
boundary. Summing over the domain gives
\begin{equation}
  \int_\Omega \phi\nabla\cdot(\rho\MM{u}) \diff x = -\int_\Omega \nabla\phi \cdot (\rho\MM{u}) \diff x + \int_\Gamma \jump{\phi\MM{u}}\tilde{\rho} \diff S, \quad \forall \phi \in \mathbb{V}^{L^2},
\end{equation}
where $\Omega$ is the domain, $\Gamma$ denotes the element facets,
$\tilde{\rho}$ denotes the value of $\rho$ on the upwind side of the
facet and the brackets $\jump{q} = q^+-q^-$ denote the jump in $q$
across the facet, where the $+$ and $-$ indicate each side of the
facet.

\subsubsection{Transport of $\mathbb{V}_\theta$ fields}
\label{subsec: theta transport}
The compressible Boussinesq and compressible Euler equation sets both
contain a prognostic equation that is the advective form of the
transport equation for a variable that is represented by a function
that is continuous in the vertical direction but discontinuous in the
horizontal direction. This means that the gradient of the field
required in the transport term is not well defined. As described in
\citet{yamazaki2017vertical}, a stable discretisation of this equation
can be constructed by using a combination of the above upwinding
strategy in the discontinuous horizontal direction and a
streamline-upwind Petrov-Galerkin (SUPG) method
\citep{brooks1982streamline} in the continuous vertical dirction. The
SUPG method replaces the finite element test function $\gamma$ with
$\gamma + \tau \MM{u}\cdot\nabla\gamma$ where $\tau$ is a
stabilisation parameter, although here we just require the vertical
part of the gradient term. This has the effect of suppressing spurious
oscillations associated with the continuous finite element
approximation of the transport term. The stabilisation parameter
$\tau$ is set to be $\Delta t/\sqrt{15}$ by default in Gusto,
following \citet{raymond1976selective}.

We first integrate by parts in each column $C$ of the mesh, taking the
upwind value of the transported field $\tilde{\theta}$ on the vertical
facets of the mesh, denoted by $\Gamma_v$, giving

\begin{align}
  \int_C \gamma\MM{u}\cdot\nabla\theta\diff x &= -\int_C \nabla\cdot(\gamma\MM{u})\theta \diff x + \int_{\Gamma_v}\gamma \MM{u}\cdot\MM{n}\tilde{\theta}\diff S, \quad \forall \gamma \in \mathbb{V}_\theta, \\
  &= \int_C \gamma\MM{u}\cdot\nabla\theta \diff x + \int_{\Gamma_v}\gamma \MM{u}\cdot\MM{n}(\tilde{\theta}-\theta_{int})\diff S, \quad \forall \gamma \in \mathbb{V}_\theta,
\end{align}
where we have integrated by parts a second time to move the
differential operator from $\gamma$ back to $\theta$ since we need to
replace the test function $\gamma$ with $\gamma+\tau w\gamma_z$, where
$w=\MM{u}\cdot\hat{\MM{z}}$ is the vertical component of the
velocity. This gives a second boundary term that now contains the
value of $\theta$ in the interior of the column, denoted by
$\theta_{int}$. Summing over the domain and applying the SUPG method
in the vertical gives
\begin{equation}
  \int_\Omega \gamma\MM{u}\cdot\nabla\theta\diff x =\int_\Omega(\gamma+\tau w\gamma_z)\MM{u}\cdot\nabla\theta \diff x + \int_{\Gamma_v}\jump{(\gamma+\tau w\gamma_z) \MM{u}}\tilde{\theta}-\jump{(\gamma+\tau w\gamma_z) \MM{u}\theta}\diff S, \quad \forall \gamma \in \mathbb{V}_\theta.
\end{equation}
Note that, as explained in \citet{cotter2023compatible}, the test
function should also be replaced in the time derivative term for
consistency.

\subsubsection{Transport of $\mathbb{V}^{HDiv}$ fields}
All three of the equation sets described in section \ref{sec:
  equations} include a prognostic equation for the velocity field
$\MM{u}\in\mathbb{V}^{HDiv}$ that contains a nonlinear transport term
$\MM{u}\cdot\nabla\MM{u}$. This term can be discretised directly using
a vector version of the scheme in section \ref{subsec: theta
  transport} or, as in \citet{natale2018variational} and
\citet{gibson2019compatible}, it can be replaced by the vector
invariant form and then discretised. In Gusto we offer both options
and we present the vector invariant scheme below.

The starting point for this discretisation is to replace the
$\MM{u}\cdot\nabla\MM{u}$ term with $(\nabla\times\MM{u})\times\MM{u}
+ \frac{1}{2}\nabla(\MM{u}\cdot\MM{u})$. Now, multiplying by a test
function and integrating by parts over an element $e$ with boundary
$\partial e$ gives
\begin{equation}
  \int_e \MM{w}\cdot(\nabla\times\MM{u})\times\MM{u}
  + \frac{1}{2}\MM{w}\cdot\nabla(\MM{u}\cdot\MM{u}) \diff x = \int_e\nabla\times(\MM{u}\times\MM{w})\cdot\MM{u}\diff x - \int_{\partial e}\MM{n}\times(\MM{u}\times\MM{w})\cdot\MM{u} \diff S - \frac{1}{2}\int_e\nabla\cdot\MM{w}\MM{u}\cdot\MM{u}\diff x, \quad \forall \MM{w} \in \mathbb{V}^{HDiv}.
\end{equation}
Sunming over the domain and taking the upwind value of $\MM{u}$,
denoted by $\tilde{\MM{u}}$, on the element boundaries gives
\begin{equation}
  \int_\Omega \MM{w}\cdot(\nabla\times\MM{u})\times\MM{u}
  + \frac{1}{2}\MM{w}\cdot\nabla(\MM{u}\cdot\MM{u}) \diff x = \int_\Omega\nabla\times(\MM{u}\times\MM{w})\cdot\MM{u}\diff x - \int_{\Gamma}\jump{\MM{n}\times(\MM{u}\times\MM{w})}\cdot\tilde{\MM{u}} \diff S - \frac{1}{2}\int_\Omega\nabla\cdot\MM{w}\MM{u}\cdot\MM{u}\diff x, \quad \forall \MM{w} \in \mathbb{V}^{HDiv}.
\end{equation}

The schemes described above do not preserve monotonicity unlike the
continuous equation, generating spurious overshoots and undershoots. When
monotonicity is important, for example, to avoid generating spurious
and unphysical negative values of moisture variables, we can apply a
limiter as part of the transport scheme. These limiters are described
in \citet{bendall2020compatible}.

It is common in weather and climate models for the horizontal and
vertical directions to be discretised differently due to the different
scales of motion in each direction. Typically the vertical part of the
transport term is treated implicitly so that the timestep is not
restricted by the vertical Courant number (e.g. as in
\citet{weller2013runge}). We provide the option to split the transport
terms into horizontal and vertical components, as described in
\citet{brown2025fast}.

@@

\authorcontribution{JS and TMB conceptualisation, investigation, methodology, project administration, software, supervision, validation and writing. JS and CJC funding acquisition, CJC conceptualisation, investigation, methodology, project administration, software, supervision, TA, AB, NH, DW software and validation, JB, TG, JH-C software.} %% this section is mandatory
